Crunching Large Graphs with Commodity Processors

### 2011-04-20 doc/hotpar2011/softxmt-hotpar2011.tex
new intro, more references

Many important emerging applications such as social network analysis,
bioinformatics, and sensor networks rely on crunching very large
graphs. Unfortunately, the computational cost of these applications
gets quickly out of hand. A Yahoo snapshot of the World Wide Web graph
from 2002 has more than 1.4 billion vertices and 6.6 billion edges
\cite{pegasus, hadi}; the social network Facebook has over 500 million
active users with 130 friends each on average
\cite{Facebook:2011p91}. Applications that process these large
datasets are slow and power-hungry. Speeding up such applications at a
low cost would have a large impact in how we deal with large data-sets
and make them even more valuable.

%For example, in January 2011, Facebook had
%500 million active users with an average of 130 friends each
%\cite{Facebook:2011p91}. Its friend suggestion algorithm in 2010 ran on
%  a 40-node cluster with 72GB memory per node, and, still, new
%  suggestions could only be computed every two days
%  \cite{Backstrom:2010p90}. Speeding up these applications at a low
%  cost would have a large impact in how we deal with large data-sets and make them even more valuable.

@@

The most interesting computational challenge comes from large,
low-diameter, power law graphs: this combination makes extracting
performance difficult. The Web graph mentioned eariler requires 120GB
of storage \cite{pegasus}, difficult to fit in a single commodity machine's
memory. It has an average diameter of 15.6 \cite{hadi}: this low diameter makes it
difficult to lay out with locality, or to partition across a
cluster. Graphs with power law distributions are also difficult to
partition in a balanced way \cite{lang, leskovec}; their few
high-degree vertices also generate much more work than the rest,
leading to load imbalances.

@@

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such an approach: it solves the memory latency problem
through concurrency rather than caching. Each XMT processor supports
128 hardware contexts and 1024 outstanding memory operations, and is
able to switch contexts every cycle \cite{tera, feo-xmt}. This ability
comes at a cost: the XMT is an expensive, non-commodity
machine.

@@

We believe we can build a system based mostly on commodity parts that
can attain XMT-like performance with a familiar, XMT-like,
programming model, but at a fraction of the cost. Therefore, our goal
is to build a system that has good performance on low-locality graph
codes but is implemented with cheap commodity processors and simple
additional support in an FPGA. This approach also has the added
benefit of not sacrificing general-purpose performance.

>>>>

Many important emerging applications such as social network analysis,
bioinformatics, and sensor networks rely on crunching very large
graphs. Unfortunately, the computational cost of these applications
gets quickly out of hand. Tools to analyze social networks and query
semantic graphs with billions of entities exist
today~\cite{joslyn:hpsf, ediger:msda, brocheler:cosi, pegasus,
  pregel}, but graphs of interest to defense applications are expected
have trillions of entities~\cite{hartley:mssg, kolda:imkd}. Speeding
up these applications at a low cost would have a large impact in how
we deal with large data-sets and make them even more valuable.

%2002 snapshot of the web has 1.4 billion vertices and 6.6 billion
%edges \cite{pegasus}. Facebook has over 500M active users who interact
%with 900M groups, events, and communities; the average user has 30 friends \cite{Facebook:2011p91}. 
%For example, in January 2011, Facebook had
%500 million active users with an average of 130 friends each
%\cite{Facebook:2011p91}. Its friend suggestion algorithm in 2010 ran on
%  a 40-node cluster with 72GB memory per node, and, still, new
%  suggestions could only be computed every two days
%  \cite{Backstrom:2010p90}. Speeding up these applications at a low
%  cost would have a large impact in how we deal with large data-sets and make them even more valuable.

% A Yahoo snapshot of the World Wide Web graph
% from 2002 has more than 1.4 billion vertices and 6.6 billion edges
% \cite{pegasus, hadi}; the social network Facebook has over 500 million
% active users with 130 friends each on average
% \cite{Facebook:2011p91}. Applications that process these large
% datasets are slow and power-hungry. Speeding up such applications at a
% low cost would have a large impact in how we deal with large data-sets
% and make them even more valuable.

@@

%  The most interesting computational challenge comes from large,
%  low-diameter, power law graphs: this combination makes extracting
% performance difficult. They do not fit in a single
% commodity machine's memory. Their low diameter means that they are
% difficult to lay out with locality: every vertex needs to be near every
% other vertex. Their power law distribution means that a few vertices
% generate much more work than the rest, leading to load
% imbalances. Together, these properties \todo{including power law distr?} lead to a
% difficult conclusion: each edge traversal is likely to lead to a very
% expensive inter-node communication.

The most interesting computational challenge comes from large,
low-diameter, power law graphs: this combination makes extracting
performance difficult. They do not fit in a single commodity machine's
memory. They are difficult lay out with locality, or to partition in a
balanced way~\cite{lang, leskovec}. Expensive inter-node communication
is a limiting factor in scalability~\cite{yoo:parallelbfs}.

% Their low diameter means that they are
% difficult to lay out with locality: every vertex needs to be near every
% other vertex. Their power law distribution means that a few vertices
% generate much more work than the rest, leading to load
% imbalances. 
% Together, these properties \todo{including power law distr?} lead to a
% difficult conclusion: each edge traversal is likely to lead to a very
% expensive inter-node communication.

% They are difficult to partition effi

% Common techniques to 
% Often

% The most interesting computational challenge comes from large,
% low-diameter, power law graphs: this combination makes extracting
% performance difficult. The Web graph mentioned eariler requires 120GB
% of storage \cite{pegasus}, difficult to fit in a single commodity machine's
% memory. It has an average diameter of 15.6 \cite{hadi}: this low diameter makes it
% difficult to lay out with locality, or to partition across a
% cluster. Graphs with power law distributions are also difficult to
% partition in a balanced way \cite{lang, leskovec}; their few
% high-degree vertices also generate much more work than the rest,
% leading to load imbalances.

@@

Multithreading is a technique that has been used successfully to
implement efficient computations for these
graphs~\cite{bader:bfsmta}. The Cray XMT is an example of such an
approach: it solves the memory latency problem through concurrency
rather than caching. Each XMT processor supports 128 hardware contexts
and 1024 outstanding memory operations, and is able to switch contexts
every cycle~\cite{tera, feo-xmt}. This ability comes at a cost: the
XMT is an expensive, non-commodity machine.

@@

We believe we can build a system based mostly on commodity parts that
can attain XMT-like performance with a familiar, XMT-like,
programming model, but at a fraction of the cost. Therefore, our goal
is to build a system that has good performance on low-locality graph
codes but is implemented with cheap commodity processors and simple
additional support in an FPGA. This approach has the added
benefit of not sacrificing general-purpose performance.
